# Additional Topics in Statistics

## Chi-Square Test ($\chi^2$-Test)
The $\chi^2$-test is used to conduct a hypothesis test on qualitative variables. Before introducing the procedure, a presentation of the $\chi^2$-distribution is necessary.

Given the values $Z_1, Z_2, \dots , Z_k$ independently drawn from standard normal distribution, then the squared sum of those values follows a $\chi^2$-distribution:
$$\sum_{i=1}^k Z_i^2 = Z \sim \chi^2_k$$
where $k$ specifies the degrees of freedom. 
```{r, echo=FALSE}
draws = 1000
k = 10
x = rep(NA,draws)
for (j in 1:draws){x[j]=sum(rnorm(10)^2)}
h = hist(x,plot=F)
ylim = range(0,h$density,0.10)
hist(x,freq=F,ylim=ylim)
curve(dchisq(x,df=k),add=TRUE)
```
To conduct the hypothesis test, consider the data on voting and education in `gss`. Using the function `CrossTable` associated with the package [gmodels](https://cran.r-project.org/web/packages/gmodels/index.html), the following table can be constructed. 
```{r}
CrossTable(gss$degree,gss$vote12,prop.r = FALSE,prop.c = FALSE,prop.chisq = FALSE,prop.t=FALSE)
```
This table is similar to what has been presented previously except that all proportions have been removed and only the counts are presented. This is a 5-by-2 contingency table (the total columns are not considered part of the table). The variable education is less than high school (0), high school (1), junior college (2), bachelor (3), and graduate (4). Assuming independence, the expected value $E$ for each cell is 
$$E=\frac{(\text{total of row})\cdot(\text{total of column})}{\text{total count}}$$
For example, consider the cell "high school" and "voting" which contains 214 counts. The expected value under the null hypothesis that voting behavior is independent of education leads to the following:
$$E_{1,1}=\frac{367 \cdot 518}{772} = 246.2513$$
Those calculations can be conducted for each cell. Then, then the $\chi^2$-test statistic is calcualted as follows:
$$\chi^2 = \sum_{i=1}^{r \cdot c} \frac{(O_i-E_i)^2}{E_i}$$
where $r \cdot c$ is number of rows multiplied by the number of colunms, $O_i$ and $E_i$ are the observed and exepected count in a cell, respectively. The degrees of freedom are calcualted as $(r-1)\cdot(c-1)$. Of course, instead of doing it manually, the following command can be used:
```{r}
chisq.test(gss$degree,gss$vote12)
```
Due to the *p*-value being very small, we reject the hypothesis of independence. The $\chi^2$ hypothesis test should only be used for qualitative data. For example, do not categorize income into quartiles to conduct a hypothesis test on whether voting depends on income. 

The example of a 2-by-2 table is a special case
```{r}
votegun   = subset(gss,
                   vote12 %in% c("voted","did not vote") &
                        owngun %in% c("yes","no"),
                   select = c("vote12","owngun"))
votegun$vote12 = ifelse(votegun$vote12=="voted",1,0)
votegun$owngun = ifelse(votegun$owngun=="yes",1,0)
t.test(votegun$vote12~votegun$owngun)
```
```{r}
chisq.test(votegun$vote12,votegun$owngun)
```





```{r}
CrossTable(votegun$vote12,votegun$owngun,prop.r = FALSE,prop.c = FALSE,prop.chisq = FALSE,prop.t=FALSE)
```

## Binomial Test
Hypothesis test about the probability of success. 

Many government forms indicate the estimated time to fill out a form. This information can be used to calculate the (financial) burden associated with providing information to government agencies. For example, the instructions of the 2020 IRS Form 1040 state that *the estimated average time burden for all taxpayers filing a Form 1040 or 1040-SR is 12 hours.* Consider the data in `irs` which contains 40 observations of taxpayers filling out the form. Conducting a regular *t*-test leads to the rejection of the hypothesis:

```{r,echo=FALSE}
t.test(irs,mean=12)
```
The hypothesis is also rejected 
```{r,echo=FALSE}
t.test(irs,mean=12,alternative="greater")
```
In the sample, 16 observations are below the estimated 12 hours and 24 observations are above 12 hours.

```{r}
binom.test(16,40)
```
checkout = c(3.8,5.3,3.5,4.5,7.2,5.1)
t.test(checkout,mu=4,alternative=c("greater"))

## Wilcoxon Signed-Rank Test
```{r,warning=FALSE,error=FALSE}
wilcox.test(irs,mu=12)
```

## Exercises
1. ***Chi-Squared Test*** (\*\*\): Use the `gss` data for this exercise. Select one of the following variables: $fulltime$, $government$, $married$, $gun$, or $deathpenalty$. Conduct a Chi-Squared hypothesis test between the variable chosen and education. Remember that the null hypothesis is independence between the variables. Report your results and interpret them as well. 