# Limited Dependent Variable Models
There are slides associated with this chapter:

- [Limited Dependent Variable Models - Slides](https://github.com/jrfdumortier/slidesstatistics/raw/main/LimitedDependentVariables.pdf)

This chapter covers three regression models in which the dependent variable is somehow limited:

1. Truncation: With truncated data, the researcher does not observe values past a particular point and those values are also not reported. Examples of truncation are low-income household studies, on-site visitation data, or time-of-use pricing experiments (excludes low-usage households). 

2. Censoring: In the case of censoring, values that are above or below a certain value are replaced by that value. For example, the demand for a particular class is not fully observed (absence of a waiting list). This is also called a Tobit model. 

3. Count dependent variable 

The following packages are necessary for this section: [AER](https://cran.r-project.org/web/packages/AER/index.html) [truncreg](https://cran.r-project.org/web/packages/truncreg/index.html), [censReg](https://cran.r-project.org/web/packages/censReg/index.html), and [pscl](https://cran.r-project.org/web/packages/pscl/index.html).

Truncation and censoring lead to a bias in the estimates. It is not always clear why or if the data is limited in its range. 



## Truncation
In the case of truncation, a certain part of the data is not observed. In the graph below, the true parameters are $\beta_0=-2$ and $\beta_1=0.5$. Values $y<0$ are not reported in the data. The green regression line is "correct" whereas the "red" is the line obtained from a regression model which ignores the truncation.  

```{r truncationplot, echo=FALSE}
bhat_real           = lm(y_real~x,data=truncation)
bhat_truncated      = lm(y_obs~x,data=truncation)
plot(truncation$x,truncation$y_obs,xlim=c(0,10),ylim=c(-3,5),pch=19,ylab="y",xlab="x")
     points(truncation$x,truncation$y_real)
     abline(bhat_real,col=c("darkgreen"))
     abline(bhat_truncated,col=c("red"))
```
If all the data was observed, the correct regression model would give the following results:
```{r truncation_bhat_real}
summary(bhat_real)
```

The estimates are biased if truncation is ignored:
```{r truncation_bhat_incorrect}
summary(bhat_truncated)

```

To correct for the truncation, use the functions from the package [truncreg](https://cran.r-project.org/web/packages/truncreg/index.html) which allows to reduce the bias of the coefficients:

```{r truncation_bhat_correct}
bhat_correct = truncreg(y_obs~x,data=truncation)
summary(bhat_correct)
```

## Censoring
In the case of censoring, the values of the dependent variable are reported at a certain point if they are above or below a certain value.

```{r, echo=FALSE}
bhat_real = lm(yreal~x,data=censoring)
bhat_censored = lm(y~x,data=censoring)
plot(censoring$x,censoring$yreal)
     points(censoring$x,censoring$y,pch=19)
     abline(bhat_real,col=c("darkgreen"))
     abline(bhat_censored,col=c("red"))
```

If all data was reported at the correct value, the following following regression model could be executed: 
```{r}
summary(bhat_real)
```

Ignoring censoring leads to biased results:
```{r}
summary(bhat_censored)
```

Using the R package [censReg](https://cran.r-project.org/web/packages/censReg/index.html)) allows for the reduction of the bias:
```{r}
b_correct = censReg(y~x,data=censoring)
summary(b_correct)
```

## Count Regression Models
Count regression models apply to cases in which the dependent variable represents discrete, integer count data. Here are some examples of count outcomes:

- What are the number of arrests for a person?
- What determines the number of credit cards a person owns? 

This section on count regression presents three models:

1. Poisson Regression Model: The condition to use this model is the absence of overdispersion, i.e., the expected value of the dependent variable is equal to the variance. 
2. Quasi-Poisson Regression Model: Overdispersion occurs if the variance of the dependent variable is larger than its mean. In this case, the Poisson Regression Model leads to unreliable hypothesis tests regarding the coefficients. The Quasi-Poisson Model solves this issue.
2. Negative Binomial Regression Model: The second possibility to deal with overdispersion is to use a Negative Binomial Regression Model. 

The main package used is [pscl](https://cran.r-project.org/web/packages/pscl/index.html). There is also an additional resource with more theoretical details on the topic: [Regression Models for Count Data in R](http://dx.doi.org/10.18637/jss.v027.i08). A more up-to-date version of the document may be found with the [pscl](https://cran.r-project.org/web/packages/pscl/index.html) package documentation. 

### Poisson Regression Model
Recall that the Poisson distribution is written as:
$$Pr(Y=k)=\frac{e^{-\lambda} \cdot \lambda^k}{k!}$$
The important characteristics of the distribution is that the mean and variance are equal to $\lambda$, i.e., $E(Y)=\lambda$ and $Var(Y)=\lambda$, this is also known as the equidispersion property. The  mean parameter is written as $\lambda=exp(\beta_0+\beta_1 \cdot x_1+ \dots + \beta_k \cdot x_k)$.

The first examples uses the 2017 [NHTS](https://nhts.ornl.gov/) data contained in `hhpub`. In a first step, the data is prepared for the regression model, i.e., missing or unknown values are eliminated and income is measured in (thousand) dollar terms:

```{r poissonregression_cleanhhpub,results=FALSE}
hhpubdata = subset(hhpub,HHFAMINC %in% c(1:11) & HOMEOWN %in% c(1,2) & URBRUR %in% c(1,2) & HHVEHCNT %in% c(0:12))
HHFAMINC = c(1:11)
INCOME = c(10,12.5,20,30,42.5,57.5,82.5,112.5,137.5,175,200)
INCOME = data.frame(HHFAMINC,INCOME)
hhpubdata = merge(hhpubdata,INCOME)
hhpubdata$RURAL = hhpubdata$URBRUR-1
hhpubdata$RENT = hhpubdata$HOMEOWN-1
```

The outcome of interest is the number of vehicles based on household income, home ownership, and urban/rural household location. Before executing the Poisson regression model, calculate the mean and variance of the outcome variable.

```{r, echo=FALSE}
mean(hhpubdata$HHVEHCNT)
var(hhpubdata$HHVEHCNT)
```
The mean and the variance are similar and thus, estimating a Poisson Regression Model is an appropriate first step. The package [AER](https://cran.r-project.org/web/packages/AER/index.html) also contains the function `dispersiontest()` which conducts a hypothesis test assuming no overdispersion. This test will be used after execution of the main regression. To estimate the model, the `glm()` function can be used by specifying `family=poisson`.

```{r }
bhat_pois = glm(HHVEHCNT~INCOME+RENT+RURAL,data=hhpubdata,family=poisson)
summary(bhat_pois)
```

All coefficients are statistically significant. The signs associated with the coefficients indicates the direction of influence on the outcome variable, i.e., the number of cars. As expected, higher income is associated with a higher number of cars and so is living in a rural environment. Renting is associated with a lower number of vehicles. Note that income and renting is possibly correlated. In general, the coefficient estimates are interpreted using $\exp(\beta)$. That is, with every unit increase in $X$, the predictor variable has a multiplicative effect of $\exp(\beta)$ on the mean of $Y$, i.e., $\lambda$:

- If $\beta=0$, then $\exp(\beta) = 1$, and the expected count $\mu= E(y) = \exp(\alpha)$, and Y and X are not related.
- If $\beta>0$, then $\exp(\beta) > 1$, and the expected count $E(y)$ is $\exp(\beta)$ times larger than when $X = 0$
- If $\beta<0$, then $\exp(\beta) < 1$, and the expected count $E(y)$ is $\exp(\beta)$ times smaller than when $X = 0$

The function `overdispersion` tests the null hypothesis of equidispersion:

```{r}
dispersiontest(bhat_pois)
```

Given the *p*-value, the null hypothesis cannot be rejected. If the data suggests overdispersion, two alternative regression models can be used: (1) Quasi-Poisson and (2) Negative Binomial.

### Quasi-Poisson Regression Model
The dataset `blm` used in this section as well as the one describing the negative binomial model is associated with the article [Black Lives Matter: Evidence that Police-Caused Deaths Predict Protest Activity](https://doi.org/10.1017/S1537592717004273). Note that the paper includes a significant number of supplementary materials which allows for the replication of the results and much more.  

The dependent variable is the total number of protests in a city. In a first step, the mean and variance of the variable $totalprotests$ are calclated:

```{r}
mean(blm$totprotests)
var(blm$totprotests)
```
The variance is significantly higher than the mean which suggests overdispersion. In a first step, a regular Poisson model is estimated.

```{r}
eq1 = "totprotests~log(pop)+log(popdensity)+percentblack+blackpovertyrate+I(blackpovertyrate^2)+percentbachelor+collegeenrollpc+demshare"
bhat1 = glm(eq1,data=blm,family=poisson)
summary(bhat1)
```
Note that the signs and statistical significance correspond to Model 1 (Table 3) in the original paper. The coefficients are different because the paper only include negative binomial regression models. The reason for not using a Poisson regression model is the presence of overdispersion:
```{r}
dispersiontest(bhat1)
```
Note that the null hypothesis is rejected at the 10\% but not the 5\% significance level. The Quasi-Poisson Regression Model handles overdispersion by adjusting the stand-errors but leaving the coefficicent estimates the same.

```{r}
bhat2 = glm(eq1,data=blm,family=quasipoisson)
summary(bhat2)
```
Note that in this case, the population density is not statistically significant anymore. 

### Negative Binomial Regression Model
The Negative Binomial Regression Model can be used in the presence of count data and overdispersion. Below, the results from the article [Black Lives Matter: Evidence that Police-Caused Deaths Predict Protest Activity](https://doi.org/10.1017/S1537592717004273) are recreated using the negative binomial models presented in the paper. The required package is [MASS](https://cran.r-project.org/web/packages/MASS/index.html)

The first model is a basic model of resource mobilization and opportunity structure.
```{r}
bhat3 = glm.nb(eq1,data=blm,link=log)
summary(bhat3) 
```

In a second model, black deaths are added:
```{r}
bhat4 = glm.nb(totprotests~log(pop)+log(popdensity)+percentblack+blackpovertyrate+I(blackpovertyrate^2)+percentbachelor+collegeenrollpc+demshare+deathsblackpc,data=blm,link=log)
summary(bhat4) 
```

And in the third model, the authors use all police-caused deaths instead (victims of any race):
```{r}
bhat5 = glm.nb(totprotests~log(pop)+log(popdensity)+percentblack+blackpovertyrate+I(blackpovertyrate^2)+percentbachelor+collegeenrollpc+demshare+deathspc,data=blm,link=log)
summary(bhat5) 
```

## Hurdle and Zero-Inflation Models
Count data often includes many observations at 0 which can lead to problems using a Poisson or a Negative-Binomial Regression Model. The application of both models is first illustrated with the `NMES1988` data from the package [AER](https://cran.r-project.org/web/packages/AER/index.html) and then with the BLM protest data. 

The data `NMES1988` contains 4406 observations of people on Medicare who are 66 years or older. The outcome of interest is the number of doctor $visits$ as a function of $hospital$ (number of hospital visits), $health$ (self-indicated health status), $chronic$ (number of chronic conditions), $gender$, $school$, and $insurance$. 

```{r}
data("NMES1988",package="AER")
eq = visits~hospital+health+chronic+gender+school+insurance
bhat_pois = glm(eq,data=NMES1988,family=poisson)
summary(bhat_pois)
bhat_nb = glm(eq,data=NMES1988)
summary(bhat_nb)
bhat_hurdle = hurdle(eq,data=NMES1988,dist="negbin")
summary(bhat_hurdle)
```

## Survival Analysis

## Exercises

1. ***Aptitude Tobit Model*** (\*\*\*): Consider the censored data set in `aptitude` in which the aptitude score is limited at 800. In a first step, estimate a regular OLS model with $apt$ as the dependent variable and $read$, $math$, and $factor(program)$ as the independent variables. In a second step, estimate a model that takes the censored nature of the data into account. Is there a significant difference in estimates?

2. ***Chicago Grocery Stores*** (\*\*\*): Subdivision of Chicago are called Chicago Community Areas (CCA). The data in `chicagogrocery` includes data about the number of grocery stores ($stores$) in each CCA as well as demographic information. Estimate a Poisson and Negative Binomial Regression Model with $stores$ as the dependent variable and the following independent variables: $income$, $pop$, unemployment rate ($unemployed/laborforce$) and percentage of blacks ($black/pop$). What do you conclude? Are the results what you would expect?

3. ***Extramarital Affairs*** (\*\*\*): Consider the data set in `fair`. The data comes from the analysis presented in the 1978 paper [A Theory of Extramarital Affairs](https://doi.org/10.1086/260646) by Ray C. Fair. The dependent variable is $affairs$ representing the number of extramarital affairs in the past year. Everything of seven and above is coded as 7. The independent variables, which we are going to use are $male$, $yearsmarried$ (number of years married), $children$, $religious$ (religiousness on a scale of 1-5 with 1 being basically an atheist), and $marriagehappiness$ (self-rating of marriage with 1=very unhappy to 5=very happy). You are going to execute five models: (1) regular OLS, (2) Probit, (3) Poisson, (4) Negative Binomial, and (5) Hurdle Model. For the Probit model, you are running a model with a binary variable of either 0 (no affair) or 1 (at least one affair). Compare the models in terms of statistical significance. What changes from one model to the next? What model is the most appropriate and why?

4. ***Biochemistry Articles*** (\*\*\*): Publish or perish summarizes life in academia. The data set `biochemistry` is based on the article [The Origins of Sex Differences in Science](https://doi.org/10.1093/sf/68.4.1297) and assesses the number of articles written by graduate students three years before receiving their Ph.D. in biochemistry. The dependent variable of interest is $articles$ and the independent variables are $female$, $married$, $kidsbelow6$, and $mentorarticles$ (number of articles by Ph.D. mentor). Estimate a quasi-poisson and a hurdle model. According to the model, what matters in terms of graduate student productivity? Why do those findings matter?

5. ***BLM II*** (\*\*\*): In a previous exercise, a regular OLS regression model was used to explain the positive number of protests. In this exercise, a zero-inflated and a hurdle model are estimated. The dependent variable for this exercise is protest frequency ($totprotests$) and the independent variables are city population ($pop$), population density ($popdensity$), percent Black ($percentblack$), black poverty rate ($blackpovertyrate$), percent of population with at least a bachelor ($percentbachelor$), college enrollment ($collegeenrollpc$), share of democrats ($demshare$), and Black police-caused deaths per 10,000 people ($deathsblackpc$). Interpret the output of the two models.

6. ***Lung Cancer*** (\*\*): The data `lung` includes the following variables: $inst$ (institution code), $time$ (survival time in days), $phecog$ (ECOG performance score as rated by the physician: 0=asymptomatic, 1=symptomatic but completely ambulatory, 2=in bed less than 50% of the day, 3=in bed more than 50% of the day but not bed bound, 4 = bed bound), $phkarno$ (Karnofsky performance score from 0=bad to 100=good as rated by physician), $patkarno$ (Karnofsky performance score as rated by the patient), $mealcal$ (calories consumed at meals) and $wtloss$ (weight loss in last six months in pounds). Plot the survival curve differentiating by sex. Estimate a survival model with $age$ and $female$ as the independent variables.  

7. ***Henning*** (\*\*\*): The data set `henning` is similar to `rossi` and comes from the paper [Cognitive-behavioral treatment of incarcerated offenders: An evaluation of the Vermont Department of Corrections' Cognitive Self-Change Program](https://doi.org/10.1177/0093854896023004001) by Henning and Frueh (1996). The following variables will be used in this analysis: $months$ (day of re-arrest measured in months), censor (dummy variable ), personal (offense against a person), property (offense against property), and cage (centered age at time of release, i.e., age minus average age).  Plot the survival curve differentiating by $personal$. Estimate three survival models with the following independent variables: (1) personal, (2) personal and property, and (3) personal, property, and cage. Interpret the output. Is there a big difference in coefficients? 



