# Probability Distributions
In order to associated more complex probabilities to a sample space, we need to employ probability distributions. In a first part, the concept of random variables is presented including expected value and variance of a random variable. The two remaining parts of this chapter cover probability distributions for discrete and continuous random variables, respectively. For each probability distribution, the respective names and commands in R are presented. Many probability distributions in R have a "name" (presented in parenthesis in the lists below). The following discrete probability distributions are presented:

- Bernoulli Distribution
- Binomial Distribution ("binom")
- Poisson Distribution ("pois")

Continuous distributions included in this chapter are:

- Uniform Distribution ("unif")
- Normal Distribution ("norm")
- Student Distribution or *t*-Distribution ("t")

There is a particular set of R commands to find various aspects of the distributions:

- `dname()` for the density or probability function
- `pname()` for the cumulative density function
- `rname()` for the random numbers 
- `qname()` for the inverse

There is also a YouTube video and slides associated with this chapter:

- [Probability Distributions - Video](https://youtu.be/29eB-nzfVAM)
- [Probability Distributions - Slides](https://github.com/jrfdumortier/slidesstatistics/blob/main/ProbabilityDistributions.pdf)

## Random Variables
A random variable $X$ is a variable that can take different values and there is a probability associated for each value or range of values. We differentiate between discrete and continuous random variables. The important aspect to keep in mind is that the sum of the probabilities for all values has to be equal to one!

### Expected Value and Variance
Think of the expected value as a weighted average. Let $X$ be a discrete random variable taking the values $x_1,x_2,\dots$  and with probability mass function $p$. Then the expected value of $X$, $E(X)$, is defined to be
$$E(X)=\sum_i x_i \cdot P(X=x_i) = \sum_i  x_i \cdot p(x_i)$$
Let $X$ be a continuous random variable taking the values with probability density function $f(x)$. Then the expected value of $X$, $E(X)$, is defined to be
$$E(X)=\int_{-\infty}^{\infty} x \cdot f(x) dx$$
The variance can be calculated in with to different equations:
$$Var(X) = E(X-E(X))^2 = E(X^2)-E(X)^2 $$

Both equations give you the variance. Sometimes one of the equations is more convenient to use. Note that $E(X^2) \neq E(X)^2$.
 
Suppose you roll a die and observe the number that comes up. The probability mass or frequency function is given by
$$p(x_i)=P(X=x_i)= \frac{1}{6} \quad \text{for i=1,...,6}$$
Thus, the expected value can be calculated as follows: 
$$E(X)  = \sum_{i=1}^6 x_i \cdot \left(\frac{1}{6}\right) = 21/6 = 3.5$$

## Discrete Probability Distributions
A random variable $X$ is said to be discrete if it can assume only a finite or countable infinite number of distinct values. A discrete random variable can be defined on both a countable or uncountable sample space. The probability that $X$ takes on the value $k$, i.e., $P(X=k)$, is defined as the sum of the probabilities of all sample points that are assigned the value k. For each value within its domain, we have $P(X=k) \geq 0$ and that the sum of all probabilities is equal to one. For example, suppose you are flipping a coin three times and associate the outcome "heads" with the number one. The results of this experiment are illustrated in the table below.

| Outcome | HHH | HHT | HTH | HTT | THH | THT | TTH | TTT |
|---------|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Pr      | 1/8 | 1/8 | 1/8 | 1/8 | 1/8 | 1/8 | 1/8 | 1/8 |
| X       |  3  |  2  |  2  |  1  |  2  |  1  |  1  |  0  |

This experiment can also be represented with a histogram which can be turned into a frequency histogram. This is illustrated for tossing a coin ten times and repeating the experiment 10,000 times. 

```{r DISrandomvariable,echo=FALSE,fig.cap="Tossing a coin ten times and repeating the experiment 1000 times."}
n = 100000  
x = rbinom(n, 10, 0.5)
t = (table(x)/n)
par(mfrow=c(1,2))
     plot(table(x),ylab="Frequency",main="n=10, p=0.5")
     plot(t,ylab="Probability",main="n=10, p=0.5") 
rm(n,x,t)
```

### Bernoulli distribution
The Bernoulli (named after Jacob Bernoulli, 1654-1705) distribution is the simplest probability distribution. There are only two outcomes: *Success* and *Failure*. The distribution is characterized by one parameter, i.e., $p$. The probability mass function is written as $P(X=1)=p$ and, correspondingly, $P(X=0)=1-p$. The expected value of the binomial is written as follows:
$$E(x)=1 \cdot p+0 \cdot (1-p)=p$$
To calculate the variance, we have $E(x^2)=1^2 \cdot p + 0^2 \cdot (1-p)=p$ and $E(x)^2=p^2$. Thus, we have 
$$Var(x)=E(x^2)-E(x)^2=p\cdot(1-p)$$ 

### Binomial distribution
The binomial distribution is closely related to the Bernoulli distribution because it represents *repeated* Bernoulli outcomes. The two parameters are $n$ and $p$. The number of successes is represented by $k$. The probability mass function is written as 
$$Pr(X=k) = {n \choose k} \cdot p^k  \cdot (1-p)^{n-k}$$
The expected value of $X$ is $E(X)=n \cdot p$ and the variance is $Var = n \cdot p \cdot (1-p)$. A situation must meet the following conditions for a random variable $X$ to have a binomial distribution: 

- You have a fixed number of trials involving a random process; let $n$ be the number of trials.
- You can classify the outcome of each trial into one of two groups: success or failure.
- The probability of success is the same for each trial. Let $p$ be the probability of success, which means $1-p$ is the probability of failure.
- The trials are independent, meaning the outcome of one trial does not influence the outcome of any other trial.

Imagine a multiple choice test with three questions and five choices for each question. First, draw a probability tree. What is the probability of two correct responses? Next, the binomial distribution is used to calculate the probability.
$$P(X=2) = {3 \choose 2} \cdot 0.2^2  \cdot (1-0.2)^{3-2} =0.096$$
and
$$P(X=3) = {3 \choose 3} \cdot 0.2^3  \cdot (1-0.2)^{3-3} =0.008$$
Summing up both probabilities gives us $P(X \ge 2) = 0.104$.

The binomial  distribution can be used to analyze the issue of overbooking. Assume that an airline as a plane with a seating capacity of 115. The ticket price for each traveler is \$400. The airline can overbook the flight, i.e., selling more than 115 tickets, but has to pay \$700 in case a person has a valid ticket but needs to be re-booked to another flight. There is a probability of 10\% that a booked passenger does not show up. The results for overbooking between 1 and 31 seats are shown in the figure below.

```{r DISoverbooking,echo=FALSE,fig.cap="Example of using the binomial distribution to determine how many seats to sell over capacity."}
seats               = 115
seatssold           = 0:30
airfare             = 400
fee                 = c(200,700,10000)
p                   = 0.9
profitfun           = function(shows,airfare,fee){airfare*pmin(shows,seats)-fee*(shows-seats)*(shows>seats)}
profit              = merge(seatssold,fee)
colnames(profit)    = c("Seats","Penalty")
profit$Profit       = 0
for(i in 1:nrow(profit)){
     x                   = p*(seats+profit$Seats[i])
     profit$Profit[i]    = profitfun(x,airfare,profit$Penalty[i])}
profit$Profit       = as.numeric(profit$Profit)
profit$Legend       = paste("Penalty fee of $",profit$Penalty,sep="")
ggplot(profit,aes(x=Seats,y=Profit,group=Legend))+geom_line(aes(color=Legend))+ylim(0,50000)+
  theme_bw()
rm(profit,airfare,fee,i,p,seats,seatssold,x,profitfun)
```

### Poisson Distribution
The Poisson distribution is used for count data (i.e., $0,1,2,...$). The probability mass function for the Poisson distribution is given by the following equation:
$$P(X=k)=\frac{\lambda^k \cdot e^{-\lambda}}{k!}$$
An example of the Poisson distribution (named after Simeon Denis Poisson, 1781-1840) for different parameter values is shown below.

```{r DISPoissonDistribution,echo=FALSE,fig.cap="Poisson Distribution"}
par(mfrow=c(3,2),mar=c(1,1,2,1),oma=c(4,4,2,0))
for (i in c(1,2,3,5,7,9)){
    curve(dpois(x,lambda=i),from=0,to=18,n=19,type="p",pch=15,cex=1.5,xlim=c(0,19),ylim=c(0,0.4),xlab="",ylab="")
    text(x=15,y=0.3,substitute(lambda==x,list(x=i)),cex=2)}
mtext("Probability Mass Function for Poisson Distribution", line = 0.5,outer=TRUE, cex = 1.2)
mtext(expression(P(X==x)),side=2,line=1.5,outer=TRUE)
mtext("X",side=1,line=1.5,outer=TRUE)
```

## Continuous distributions
We define a random variable to be continuous if $F_X(x)$ is a continuous function of $x$ and for every real number $x$. The probability density function, $f_X(x)$, of a continuous random variable X is the function $f(\cdot)$ that satisfies
$$F_X(x)=\int_{-\infty }^{x} f_{X}(u) du$$
Properties are that $f_X(x) \geq 0$ for all $x$ and 
$$ \int_{-\infty }^{\infty } f_X(x)  dx =1$$
It is important to note that for a continuous distribution, the probability of a particular event is zero.

### Uniform distribution
The simplest continuous distribution is the uniform distribution. If $a<b$, a random variable $X$ is said to have a uniform probability distribution on the interval $(a,b)$ if and only if the density function of $X$ is
$$f(x) = \frac{1}{b-a}$$
Parameters defining a uniform distribution: $a,b$. Suppose that a package is scheduled to be delivered between 9:00 (9am) and 20:00 (8pm). You are out for lunch between 12:00 (noon) and 13:30 (1:30pm). The probability that the package arrives during the lunch break is $P(12<x<13)=\frac{1}{20-9} \cdot 1.5 \approx 13.64\%$.
 
### Normal distribution
The random variable $X$ is said to be normally distributed with mean $\mu$ and variance $\sigma^2$ (abbreviated by $x \sim N[\mu,\sigma^2$) if the density function of $x$ is given by
$$f(x;\mu,\sigma^2)=\frac{1}{\sqrt{2\pi \sigma^2}}\cdot e^{{\frac{-1}{~2} \left(\frac{x-\mu}{\sigma}\right)}^2}$$
The Normal distribution can be derived from the Binomial Distribution. 

```{r DISbinomialtonormal,echo=FALSE}
par(mfrow=c(2,3))
  n = 50
  p = 0.5
  d = c(10,100,1000,10000,100000,1000000)
  for(i in 1:6){
    x = rbinom(d[i],n,p)
    hist(x,prob=TRUE,col="grey",ylim=c(0,0.2),xlim=c(10,40),main=paste(d[i]," Draws"))
    curve(dnorm(x,n*p,sqrt(n*p*(1-p))),add=TRUE,col="red",lwd=2)}
```

```{r,echo=FALSE}
par(mfrow=c(1,2))
          curve(dnorm(x,mean=0,sd=1),from=-4,to=4,ylab="Density",xlab="x")
               x=seq(-4,-1.5,length=100)
               y=dnorm(x)
               polygon(c(-4,x,-1.5),c(0,y,0),col="gray")
          curve(dnorm(x,mean=0,sd=1),from=-4,to=4,ylab="Density",xlab="x")
               x=seq(-1.5,-0.5,length=100)
               y=dnorm(x)
               polygon(c(-1.5,x,-0.5),c(0,y,0),col="gray")
```

The normal probability density function is bell-shaped and symmetric as shown in the figure. It is usually necessary to standardize the normal distribution to make it N(0,1):
$$z = \frac{X-\mu}{\sigma}$$
where $z$ is the distance from the mean of a Normal distribution expressed in units of the standard deviation. Suppose that we have $\mu = 75$ and $\sigma = 10$ and we want to find $P(x<60)$ and $P(60<x<70)$. 
$$Pr(x<60) \Rightarrow \frac{60-75}{10}=-1.5\\
  P(60<x<70) \Rightarrow z_1 = -1.5, z_2 = -0.5$$
The normal distribution can be derived from the binomial distribution (Galton board). Assume that the average height of males and females in the U.S. is $\mu_M=70$ and $\mu_F=65$ inches, respectively. The standard deviations are $\sigma_M=4$ and $\sigma_F=3.5$. Calculate the probabilities associated with being $\sigma$, $2 \cdot \sigma$, and $3 \cdot \sigma$ away from the mean. For more information on how the Normal Distribution evolved, see [The Evolution of the Normal Distribution](https://doi.org/10.1080/0025570X.2006.11953386). 

### *t*-Distribution
Hence, we have to replace it with an estimate, i.e., with the value of the sample standard deviation. To calculate the Student $t$ distribution, we need the degrees of freedom as well: $t_{\alpha,\nu}$. This distribution was published by William Gosset in 1908. His employer, Guinness Breweries, required him to publish under a pseudonym, so he chose "Student".  

```{r DIStdistribution,echo=FALSE,fig.cap="Student Distribution for various levels of degrees of freedom and Normal distribution"}
x       = seq(-4, 4, length=100)
hx      = dnorm(x)
degf    = c(1, 3, 8, 30)
colors  = c("red", "blue", "darkgreen", "gold", "black")
labels  = c("df=1", "df=3", "df=8", "df=30", "normal")
plot(x,hx,type="l",lty=2,xlab="x value",ylab="Density",main="Comparison of t Distributions")
for (i in 1:4){lines(x, dt(x,degf[i]), lwd=2, col=colors[i])}
legend("topright", inset=.05, title="Distributions",
  labels, lwd=2, lty=c(1, 1, 1, 1, 2), col=colors)
```

## Distribution Fitting
Bins of a histogram:
     \begin{itemize}
\item Bin width: $B_1,B_2,\dots,B_m$
     \item Bin height: Number of observations in the bin / $(n \cdot |B_i|)$
     \end{itemize}



For example, consider the multiple choice questionnaire with three questions and five possible answers. What is the probability of having all three correct? What is the probability of having at most 2 correct?
     
     <<DIS_Binomial,echo=TRUE,message=FALSE,warning=FALSE,error=FALSE,include=TRUE,fig.keep="none">>=
     dbinom(3,3,1/5)  
pbinom(2,3,1/5)
pbinom(4,14,0.3)
qnorm(0.8,21.1,5.1)
@
     

## Exercises
1. *Fair Game*: A friend of yours has a coin and proposes the following game. You toss the coin 10 times and count the number of heads. The amount you win or lose on $k$ heads is given by $k^2 - 7\cdot k$. 
     (a) Plot the payoff function.
     (b) Make an exact computation using R to decide if this is a good bet.
     (c) Use the function `rbinom()` and generate 100,000 random outcomes of the game. Calculated the expected winning and compare it to your calculation in part (b).

2. *O'Neill Air*: The probability of any O'Neill Air flight being delayed more than 15 minutes is 0.1. We randomly select four different O'Neill Air flights.
     (a) Calculate the probability that all four flights arrived within 15 minutes of the scheduled time?
     (b) Calculate the probability that none of the selected flights arrived within 15 minutes of the scheduled time?
     (c) Calculate the probability that at least one of the selected flights arrived within 15 minutes of the scheduled time?

3. *Apple Juice Machine*: An apple juice company wants to purchase a new bottling machines to fill 16 ounce cans. Two manufactures indicate the following performances for their machines. The first machine fills the cans with 16.5 ounces and a standard deviation of 0.3. The second machine fills the cans with 16.2 ounces and standard deviation 0.1. The filling quantities for both machines are normally distributed. Anything below 16 ounces cannot be sold since it does not meet the advertised 16 ounces. Which machine does the apple juice producer need to buy in order the minimize the quantity of cans which cannot be sold. 

4. *ACT*: Assume ACT scores are normally distributes with mean 19 and standard deviation of 4. A university accepts applicants which score in the top 20\%. What is the minimum ACT score that gets you accepted? Calculate the probability of 0, 1, \dots , 10 students meetings the requirements out of 10 students.

5. *LED Light Bulbs*: On average, a LED light bulb manufacturer produces 250 defective light bulbs out of 5,000. You randomly pick 8 light bulbs. What is the probability of finding more than 1 defective bulbs?

6. *Vaccine Trials*: A pharmaceutical company estimates that a new vaccine that is undergoing test trials has a 1/10000 chance of causing a serious side effect in a human being. If the company administers the vaccine to 10,000 volunteers, what is the probability that at least 1 volunteers develops those serious side effects? The company wants to make sure that the probability of at least one person developing the side effect is 95\%. How many volunteers do they need?

7. *Quillayute*: Airline magazines are often not very interesting. On a flight from Washington DC to Indianapolis, a magazine did cover an interesting story however. A resort in WA charges the temperature in Dollars per night. For example, if it is 72F then a room costs \$72. Of course, this is too boost demand in the Winter months. The closest weather station to the resort is at Quillayute Airport.