# Qualitative Choice Models

The last chapter introduced binary choice models that answer questions such as:

- Did you vote during the last election?
- Does an individual recidivate after being released from prison?
   
In this chapter, cases of dependent variables with more than two categories are considered. The categorical dependent variables can be with or without natural ordering. Here are some examples of natural ordering:

- Level of happiness: Very happy, happy, okay, or sad
- Intent to get vaccinated: Definitely yes, probably yes, probably no, definitely no

Examples without natural ordering:

- Type of car to purchase: Passenger car, Pick-up truck, van, convertible

Those models are also known as Discrete Choice Models. The packages required for this section are [mlogit](https://cran.r-project.org/web/packages/mlogit/index.html), [erer](https://cran.r-project.org/web/packages/erer/index.html), [MASS](https://cran.r-project.org/web/packages/MASS/index.html), [AER](https://cran.r-project.org/web/packages/AER/index.html), [glmmML](https://cran.r-project.org/web/packages/glmmML/index.html), and [nnet](https://cran.r-project.org/web/packages/nnet/index.html). The data required are `fpdata` and `evdata` in addition to `Fishing` and `TravelMode` which are part of the packages [mlogit](https://cran.r-project.org/web/packages/mlogit/index.html) and [AER](https://cran.r-project.org/web/packages/AER/index.html). You can load them into R with the following commands:

- `data("Fishing",package="mlogit")`
- `data("TravelMode",package="AER")`

There is also a [YouTube Video](https://youtu.be/xLUIil1sGW4) associated with this chapter. 

## Ordered Logit Model
Suppose that respondents are asked if they are going to get a COVID-19 vaccine. The choices are 

- Definitely yes
- Probably yes
- Probably no
- Definitely no

The ordered logit model assumes the presence of a latent (unobserved by the researcher) variable $y^*$:
$$y_i^* = \beta_0+\beta_1 \cdot x_i + \epsilon_i$$
In the case of the vaccine model, this may be a measure of trust in vaccines. What the researcher does measure is an $m$-alternative ordered model:
$$y_i = j \quad \text{if} \quad \alpha_{j-1} < y_i^* \leq \alpha_j \quad \text{for} \quad  j=1, \dots,m$$
where $\alpha_0 = - \infty$ and $\alpha_m=\infty$. In this case, we have
$$ \begin{align*}
    Pr(y = j) &=Pr(\alpha_{j-1} < y_i^* \leq \alpha_j)\\
                &=Pr(\alpha_{j-1} < \beta_0+\beta_1 \cdot x_i + \epsilon_i \leq \alpha_j)\\
                &=Pr(\alpha_{j-1}- \beta_0-\beta_1 \cdot x_i < \epsilon_i \leq \alpha_j - \beta_0-\beta_1 \cdot x_i)\\
                        &=F(\alpha_j - \beta_0-\beta_1 \cdot x_i)-F(\alpha_{j-1} - \beta_0-\beta_1 \cdot x_i)
    \end{align*}$$
For the ordered logit: $F(z)=exp(z)/(1+exp(z))$. The cut-off example points for the graph below are $\alpha_0 = -\infty$, $\alpha_2 = -1.5$, $\alpha_3 = 2$, and $\alpha_4 = \infty$

```{r ordered_logit_plot, echo=FALSE}
logit_cdf           = function(z){exp(z)/(1+exp(z))}
alpha1              = -1.5
alpha2              = 2
curve(logit_cdf,from=-5,to=5,yaxs="i",xaxs="i")
segments(alpha1,0,alpha1,logit_cdf(alpha1),yaxs="i",xaxs="i",lty=2)
segments(alpha2,0,alpha2,logit_cdf(alpha2),yaxs="i",xaxs="i",lty=2)
```

### Ordered Logit Example: Organic Food Purchase
The order logit model is illustrated with a survey on the purchase frequency of organic tomatoes and organic strawberries `fpdata`:

- Never (1), rarely (2), once per month (3), every 2 weeks (4), 1-2 times a week (5), almost daily (6)

The independent variables included in the model are

- Age and female
- Education: High school (1), some college (2), bachelor (3), master (4), technical school diploma (5), doctorate (6)

```{r}
fpdata$strawberriesorg = as.factor(fpdata$strawberriesorg)
strawdata = fpdata[c("strawberriesorg","age","education","female","kidsunder12")]
strawdata = na.omit(strawdata)
bhat = polr(strawberriesorg~age+education+female+kidsunder12,data=strawdata,Hess=TRUE)
summary(bhat)
```
For the organic purchases data, the cut-off points are under "Intercepts" and thus, we have (rounded coefficients):
$$z =  -0.020 \cdot age  +0.0160 \cdot education -0.415 \cdot female +0.286 \cdot kidsunder12$$
The cut-off points can be interpreted as follows:
$$\begin{align*}
    Pr(y=1) &= P(z+\epsilon_i \leq -1.4958)\\
    Pr(y=2) &= P(-1.4958 < z+\epsilon_i \leq -0.4381)\\
    Pr(y=3) &= P(-0.4381 < z+\epsilon_i \leq  0.2084)\\
    Pr(y=4) &= P(0.2084 < z+\epsilon_i \leq 0.8352)\\
    Pr(y=4) &= P(0.8352 < z+\epsilon_i \leq 1.6314)\\
    Pr(y=6) &= P(1.6314 \leq z+\epsilon_i )
\end{align*}$$
In order to get the $p$-values displayed in the output, you have to execute two additional steps:

```{r}
bhat.coef = data.frame(coef(summary(bhat)))
bhat.coef$pval = round((pnorm(abs(bhat.coef$t.value),lower.tail=FALSE)*2),2)
bhat.coef
```

### Predicted Probability and Marginal Effects
The predicted probability for each observation can be obtained as well assuming that the output of the `polr` command is stored in $bhat$.

```{r}
bhat.pred = predict(bhat,type="probs")
x = ocME(bhat)
x$out$ME.all
```

## Multinomial Logit and Multinomial Probit Models
Revealed preferences:

- Observed choices of individuals

Stated preference

- Hypothetical choice situations

Economists' modeling of choice

- Utility/happiness/satisfaction associated with multiple choice situations

### Theoretical Aspects
Travel choice model dependent on cost ($x$) and time ($z$):
$$V_j = \alpha_j + \beta_1 \cdot x_j + \beta_2 \cdot z_j$$
Probability of choosing alternative $j$ (assuming three choices):
$$\begin{align*}
    P(1) &= \frac{e^{V_1}}{e^{V_1}+e^{V_2}+e^{V_3}}\\
    P(2) &= \frac{e^{V_2}}{e^{V_1}+e^{V_2}+e^{V_3}}\\
    P(3) &= \frac{e^{V_3}}{e^{V_1}+e^{V_2}+e^{V_3}}
\end{align*}$$
Note that $P(1)+P(2)+P(3) = 1$

### Data Managment
Long shape

- One row for each alternative

Wide shape

- One row for each choice situation

There are some very good resources on data management and the package in general:

- [Estimation of Random Utility Models in R: The mlogit Package](http://dx.doi.org/10.18637/jss.v095.i11)

Travel Mode (long format)

- [Travel Mode Choice Data](https://rdrr.io/cran/AER/man/TravelMode.html)
- `mlogit.data(travelmode,choice="choice",shape="long",alt.levels=c("air","train","bus","car"))`

### Fishing Data
The data is in wide format:

- Fishing modes: beach, pier, private, and charter
- Alternative-specific variables: price and catch
- Individual-specific variables: income
- Suitability of the "wide" format to store individual-specific variables
- The R parameter `varying` designates alternative specific variables

In a first step, the model is only estimated using the individual-specific variable $income$:

```{r}
data("Fishing",package="mlogit")
fishing = mlogit.data(Fishing,shape="wide",varying=2:9,choice="mode")
bhat = mlogit(mode~0|income,fishing)
summary(bhat)
fishing.fitted      = fitted(bhat,outcome=FALSE)
effects(bhat,covariate ="income")
```

In a second step, alternative-specific variables are added:

```{r}
bhat                = mlogit(mode~catch+price|income,data=fishing)
summary(bhat)
fishing.fitted      = fitted(bhat,outcome=FALSE)
effects(bhat,covariate="income")
rm(bhat,Fishing,fishing.fitted)
```

The `mlogit` package also allows for the estimation of a multinomial probit model.
```{r}
bhat                = mlogit(mode~catch+price|income,data=fishing,probit=FALSE)
summary(bhat)
fishing.fitted      = fitted(bhat,outcome=FALSE)
effects(bhat,covariate="income")
rm(bhat,Fishing,fishing,fishing.fitted)
```

### Travel Data
```{r}
data("TravelMode",package="AER")
travelmode = mlogit.data(TravelMode,choice="choice",shape="long",alt.var="mode")
bhat = mlogit(choice~gcost+wait|income+size,data=travelmode,reflevel="car")
summary(bhat)
tavel.fitted = fitted(bhat,outcome=FALSE)
```

### Electric Vehicle Data
```{r}
evdata = mlogit.data(evdata,shape="wide",choice="choice")
bhat = mlogit(choice~0|age+female+level2+numcars+edu+income+politics,data=evdata)
summary(bhat)
evdata.fitted = fitted(bhat,outcome=FALSE)
rm(bhat,evdata,evdata.fitted)
```

## Exercises

1. ***Happiness*** (\*\*\*): Using the data set `happy`, generate an ordered logit regression model that regresses the dependent variable $happiness$ on those variables that have the strongest potential causal relationship. For your model, interpret the R output and indicate why each independent variable that is included in the model would contribute to higher or lower happiness. Speak to the possibility of multicollinearity in the independent variables.

2. ***AFV Choice*** (\*\*\*): The data `evdata` contains data about the choice of consumers with respect to alternative fuel vehicles. For each consumer, you have the following variables: $age$, $suv$ (whether they are interested in buying a SUV), $level2$ (indicating whether people have a fast charger for electric cars in their community), $own \dots$ (indicating whether the respondent currently has a gas, hybrid, plug- in hybrid, or battery electric vehicle), $gender$ (1=female) and $numcars$ (number of cars). Estimate a multinomial logit model that estimates the probability of a consumer to purchase a gasoline, hybrid, plug-in hybrid, or battery electric vehicle. Calculate the marginal probabilities as well.

3. ***NHTS*** (\*\*\*): Consider the data set `hhpub` from the 2017 National Household Travel Survey (NHTS). The data contains information about household characteristics and some of their travel means. For this question, you will focus on the following variables: $BIKE$, $HBPPOPDN$, $HHFAMINC$, $HHVEHCNT$, $HOMEOWN$, and $URBRUR$. You must read the [codebook](https://nhts.ornl.gov/tables09/CodebookBrowser.aspx) for this question and learn how the variables are coded. Go to the [codebook](https://nhts.ornl.gov/tables09/CodebookBrowser.aspx) and pick “Household” as the dataset (drop down menu). Before conducting the analysis, delete all entries
that are not complete (i.e., all the negative values). Once you have final data set, estimate
an ordered logit model with $BIKE$ as the dependent variable and the other variables as the
independent variables. What do you conclude from the model?

4. ***Home Heating*** (\*\*\*): Consider the dataset `Heating` from the  You can load the data set into R by typing: `data("Heating",package="mlogit")`. The data contains the choice of heating systems in California homes. 
    
    Estimate a multinomial logit model with installation and operating cost as the alternative-specific variables and income, age, and number of rooms as the individual-specific variables.
