# Simultanous Equation Models

http://rstudio-pubs-static.s3.amazonaws.com/195495_d47be7c818424a3c9bfa5452dd17c6c6.html

The models presented in the previous chapters contained one dependent ($y$) and multiple independent variables ($x_k$). The relationship was such that variables $x_k$ influence $y$. 

The package required for this chapter is called [systemfit](https://cran.r-project.org/web/packages/systemfit/index.html). 

## Examples
Consider the relationship between law enforcement size and crime. Often the relationship is considers to be as follows:
$$crimerate = \beta_{1,0} + \beta_{1,1} \cdot le + \beta_{1,k} \cdot x_k$$
But at the same time, it can be argued that the relationship is:
$$le = \beta_{2,0} + \beta_{2,1} \cdot crimerate + \beta_{2,k} \cdot x_k$$

In economics, the most well-known example is the joint determination of price and quantity in a supply and demand model. 

Consider a simple supply and demand model which can be written as follows:

$$ Q^D = \alpha_0 + \alpha_1 \cdot P + \alpha_2 \cdot I + \epsilon_D\\
   Q^S = \beta_0 + \beta_1 \cdot P + \epsilon_S$$

Only equilibrium values, i.e., $Q^D=Q^S$, are observed and thus, the equilibrium conditions are as follows:

$$ Q = \frac{\alpha_1 \cdot \beta_0-\alpha_0 \cdot \beta_1}{\alpha_1-\beta_1}-\frac{\alpha_2 \cdot \beta_1}{\alpha_1 \cdot \beta_1} \cdot I + \frac{-\beta_1 \cdot \epsilon^D + \alpha_1 \cdot \epsilon^S}{\alpha_1-\beta_1}\\
P=\frac{\beta_0-\alpha_0}{\alpha_1-\beta_1}-\frac{\alpha_2}{\alpha_1-\beta_1} \cdot I+\frac{\epsilon^S-\epsilon^d}{\alpha_1-\beta_1}$$

The above equation is called a *reduced form* equation. Both equations are necessary to determine $P$ and $Q$. Moreover, both equations include the error term from the demand and supply equation. Use of the regular OLS Model yields biased results. 

The supply and demand model consider above also reveals the problem of identification. Assume having data on $P$, $Q$, and $I$. And define $\gamma_i$ as follows:
$$ \gamma_1 = \frac{\alpha_1 \cdot \beta_0-\alpha_0 \cdot \beta_1}{\alpha_1-\beta_1}\\
\gamma_2 = \frac{\alpha_2 \cdot \beta_1}{\alpha_1 \cdot \beta_1}\\
\gamma_3 = \frac{\beta_0-\alpha_0}{\alpha_1-\beta_1}\\
\gamma_4 = \frac{\alpha_2}{\alpha_1-\beta_1}$$

You can estimate the $\gamma_i$ coefficients but there are total of five unknowns. Thus, it is not possible to recover the coefficients from the original supply and demand model.  

## Application in R
The first application replicates a food market example found in [Elements of Econometrics](https://doi.org/10.3998/mpub.15701) by Jan Kmenta. The data is included in the package [systemfit](https://cran.r-project.org/web/packages/systemfit/index.html). The demand side of the model is written as:
$$consump = \beta_{1,0} + \beta_{1,2} \cdot price + \beta_{1,3} \cdot income$$
The supply side of the model is written as:
$$consump = \beta_{2,0} + \beta_{2,2} \cdot price + \beta_{2,3} \cdot farmPrice + \beta_{2,4} \cdot trend$$

```{r}
data("Kmenta")
eqDemand = consump~price+income
eqSupply = consump~price+farmPrice+trend
eqSystem = list(demand=eqDemand,supply=eqSupply)
bhat_seq = systemfit(eqSystem,data=Kmenta,method="OLS")
summary(bhat_seq)
```


### Klein's Model
Consumption
```{r}
data("KleinI")
eqConsump = consump~corpProf+corpProfLag+wages
eqInvest = invest~corpProf+corpProfLag+capitalLag
eqPrivWage = privWage~gnp+gnpLag+trend
#inst =- ~ govExp + taxes + govWage + trend + capitalLag + corpProfLag + gnpLag
system =list(Consumption=eqConsump,Investment=eqInvest,PrivateWages=eqPrivWage )
```

```{r}
summary(systemfit(system,data=KleinI))

```



```{r}
bhat = lm(C~I,data=simeqexample)
summary(bhat)
```

```{r}
demandeq = q~p+rdi
supplyeq = q~p
bhat = lm(demandeq,data=Book1)
summary(bhat)
bhat = lm(supplyeq,data=Book1)
summary(bhat)
eqSystem = list(demand=demandeq,supply=supplyeq)
bhat_seq = systemfit(eqSystem,data=Book1,method="SUR")
summary(bhat_seq)

```


